---
title: "reph mate guarding"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: 
  html:
    code-fold: true
    code-tools: true
    self-contained: true
    toc: true
    fig-height: 6
execute:
  warning: false
---

## DATA, SETTINGS & PACKAGES

```{r, settings_data}
#| echo: true

  sapply(c(
    "data.table", 
    "glue", 
    "lubridate",
    "glmmTMB", 
    "ggplot2",
    "DHARMa",
    "performance",
    "effects"
  ),
  require,
  character.only = TRUE
  )

  dt2hh <- function(x) {
    h <- as.POSIXlt(x)
    h$hour + h$min / 60 + h$sec / 3600
    }

  hh2rad <- function(x) {
    x * pi / 12
    }

# DATA
  dp <- fread( here::here("DATA/PAIR_WISE_INTERACTIONS_BREEDING_PAIRS.txt"),
    nThread = 20, tz = "UTC")
  setorder(dp, pairID, datetime_1)

  dp[, datetime_round := round_date(datetime_1, unit = "3 hours")]

  dp[, period := fcase(
    datetime_rel_pair0 >= -5 & datetime_rel_pair0 <= -1, "[-5,-1]", 
    datetime_rel_pair0 >= 0 & datetime_rel_pair0 <= 3,   "[0,3]", 
    datetime_rel_pair0 >= 4 & datetime_rel_pair0 <= 10,  "[4,10]"
  )]

  X = dp[!is.na(period), .(prop_interactions = mean(interaction), 
            datetime_rel_pair  = mean(datetime_rel_pair), 
            datetime_rel_pair0 = min(datetime_rel_pair0),
            initiation_rel = mean(initiation_rel0)
            
            ), 
        by = .(nestID, year_, date_,period, datetime_round)]

  X[, HH := dt2hh(datetime_round)]

  # beta models only accept proportion in the (0,1) interval.
  X[prop_interactions == 1, prop_interactions := 0.9999]
  X[prop_interactions == 0, prop_interactions := 0.0001]

  # mean by nestID for initiation_rel plot
  xp = dp[, .(prop_interactions = mean(interaction), .N), .(nestID, initiation_rel, period)]

```

## METHODS

By utilizing the interaction at lag1  as an autocovariate, we discovered that temporal autocorrelation occurs mostly at lag1 and lag2. 
Therefore data aggregation (e.g. combining the 10 mins blocks into 3 hours blocks) should take care of the AC in the model.
We aggregate the raw interaction data (1/0) into proportions. 

We'll use [DHARMa](https://cran.r-project.org/web/packages/DHARMa/vignettes/DHARMa.html) by  Florian Hartig and [performance](https://easystats.github.io/performance/index.html) by Daniel Lüdecke for model fitting evaluation. 

The histograms below suggest a Gaussian model will probably perform poorly but a beta distribution glmm would work well. Indeed, when looking at posterior predictive plots we see that the lines predicted by the `beta` models follow the observed data reasonably well while the lines predicted from a Gaussian model don't. 

__TODO__ @Hannes: `initiation_rel` should not be rounded by day. 

```{r}
ggplot(X, aes(x = prop_interactions)) + 
facet_wrap(~period, ncol = 1, scale = "free") + 
geom_histogram()


```


## Period `[-5,-1]`

### Beta glmm

```{r}
  dx <- X[period == "[-5,-1]"]
  setorder(dx, nestID, datetime_rel_pair)
  dx[, time := 1:.N, .(date_, nestID)]

  fm <- glmmTMB(
    prop_interactions ~
      sin(hh2rad(HH)) + 
      cos(hh2rad(HH)) + 
      poly(initiation_rel, 2) + 
      poly(datetime_rel_pair, 2) +
      (1 + datetime_rel_pair + I(datetime_rel_pair^2) | nestID) 
      ,
      family = beta_family(link="logit"), 
    data = dx,
    control = glmmTMBControl(parallel = 30)
  )

  summary(fm)

```

### Model checks



```{r}
posterior_predictive_check(fm)

simulateResiduals(fm) |>
recalculateResiduals(group = dx$time) |>
testTemporalAutocorrelation(time = unique(dx$time))

```

Do we really need a `beta` model? We can try fit a Gaussian model and see if it fits the data as good. 

```{r}
  fmg <- glmmTMB(
    prop_interactions ~
      sin(hh2rad(HH)) + 
      cos(hh2rad(HH)) + 
      poly(initiation_rel, 2) + 
      poly(datetime_rel_pair, 2) +
      (1 + datetime_rel_pair + I(datetime_rel_pair^2) | nestID) 
      ,
    data = dx,
    control = glmmTMBControl(parallel = 30)
  )

  posterior_predictive_check(fmg)

```

The plot above suggest that the gaussian model is a poor choice for our data. 

### Effect plots

Relative date effects 
```{r}
  EE <- allEffects(fm, xlevels = 100)

  e <- EE$"poly(datetime_rel_pair,2)"
  ed <- e |>
    data.frame() |>
    setDT()

  ggplot(ed, aes(y = fit, x = datetime_rel_pair)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.5) + 
    geom_line(linewidth = 2) 

  EE <- allEffects(fm, xlevels = 100)

  e <- EE$"poly(initiation_rel,2)"
  ed <- e |>
    data.frame() |>
    setDT()

  ggplot(ed, aes(y = fit, x = initiation_rel)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.5) + 
    geom_line(linewidth = 2) + 
    geom_point(data = xp[period == "[-5,-1]"], aes(y = prop_interactions, size = N), shape = 21) 

```

Circadian effects back-transformed to time of day from the trigonometric scale. 
```{r}
  EE <- allEffects(fm, xlevels = 100)

  e <- EE$"sin(hh2rad(HH))"
  ed <- e |>
    data.frame() |>
    setDT()

  ggplot(ed, aes(y = fit, x = HH)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.5) + 
    geom_line(linewidth = 1) +
     scale_y_continuous(limits = c(-0.01, 1.01), breaks = seq(0, 1, 0.1), 
                     labels = c('0.0', '', '0.2', '', '0.4', '', '0.6', '', '0.8', '', '1.0'),
                     expand = expansion(add = c(0, 0.05))) +
    xlab("Hour") + 
    ylab("% interactions")
```






## Period `[0,3]`
### Beta glmm

```{r}
  dx <- X[period == "[0,3]"]
  setorder(dx, nestID, datetime_rel_pair)
  dx[, time := 1:.N, .(date_, nestID)]


  fm <- glmmTMB(
    prop_interactions ~
      sin(hh2rad(HH)) + 
      cos(hh2rad(HH)) + 
      poly(initiation_rel, 2) + 
      poly(datetime_rel_pair, 2) +
      (1 + datetime_rel_pair + I(datetime_rel_pair^2) | nestID)
      ,
      family = beta_family(link="logit"), 
    data = dx,
    control = glmmTMBControl(parallel = 30)
  )

  summary(fm)


```
### Model checks



```{r}
check_collinearity(fm) 
posterior_predictive_check(fm)
simulateResiduals(fm) |>
recalculateResiduals(group = dx$time) |>
testTemporalAutocorrelation(time = unique(dx$time))

```



### Effect plots

Relative date effects 
```{r}
  EE <- allEffects(fm, xlevels = 100)

  e <- EE$"poly(datetime_rel_pair,2)"
  ed <- e |>
    data.frame() |>
    setDT()

  ggplot(ed, aes(y = fit, x = datetime_rel_pair)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.5) + 
    geom_line(linewidth = 2)



  EE <- allEffects(fm, xlevels = 100)

  e <- EE$"poly(initiation_rel,2)"
  ed <- e |>
    data.frame() |>
    setDT()

  ggplot(ed, aes(y = fit, x = initiation_rel)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.5) + 
    geom_line(linewidth = 2) + 
    geom_point(data = xp[period == "[0,3]"], aes(y = prop_interactions, size = N), shape = 21)




```


Circadian effects back-transformed to time of day from the trigonometric scale. 
```{r}
  EE <- allEffects(fm, xlevels = 100)

  e <- EE$"sin(hh2rad(HH))"
  ed <- e |>
    data.frame() |>
    setDT()

  ggplot(ed, aes(y = fit, x = HH)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.5) + 
    geom_line(linewidth = 1) +
    scale_y_continuous(limits = c(-0.01, 1.01), breaks = seq(0, 1, 0.1), 
                     labels = c('0.0', '', '0.2', '', '0.4', '', '0.6', '', '0.8', '', '1.0'),
                     expand = expansion(add = c(0, 0.05))) +
    xlab("Hour") + 
    ylab("% interactions")



```

## Period `[4,10]`

### Beta glmm

```{r}
  dx <- X[period == "[4,10]"]
  setorder(dx, nestID, datetime_rel_pair)
  dx[, time := 1:.N, .(date_, nestID)]


  fm <- glmmTMB(
    prop_interactions ~
      sin(hh2rad(HH)) + 
      cos(hh2rad(HH)) + 
      poly(initiation_rel, 2) + 
      poly(datetime_rel_pair, 2) +
      (1 + datetime_rel_pair  | nestID)
      ,
      family = beta_family(link="logit"), 
    data = dx,
    control = glmmTMBControl(parallel = 30)
  )

  summary(fm)





```
### Model checks



```{r}
check_collinearity(fm) 
posterior_predictive_check(fm)
simulateResiduals(fm) |>
recalculateResiduals(group = dx$time) |>
testTemporalAutocorrelation(time = unique(dx$time))

```

### Effect plots

Relative date effects 
```{r}
  EE <- allEffects(fm, xlevels = 100)

  e <- EE$"poly(datetime_rel_pair,2)"
  ed <- e |>
    data.frame() |>
    setDT()

  ggplot(ed, aes(y = fit, x = datetime_rel_pair)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.5) + 
    geom_line(linewidth = 2) 

  EE <- allEffects(fm, xlevels = 100)

  e <- EE$"poly(initiation_rel,2)"
  ed <- e |>
    data.frame() |>
    setDT()

  ggplot(ed, aes(y = fit, x = initiation_rel)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.5) + 
    geom_line(linewidth = 2) + 
    geom_point(data = xp[period == "[4,10]"], aes(y = prop_interactions, size = N), shape = 21)

```

Circadian effects back-transformed to time of day from the trigonometric scale. 
```{r}
  EE <- allEffects(fm, xlevels = 100)

  e <- EE$"sin(hh2rad(HH))"
  ed <- e |>
    data.frame() |>
    setDT()

  ggplot(ed, aes(y = fit, x = HH)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.5) + 
    geom_line(linewidth = 1) +
    xlab("Hour") + 
    ylab("% interactions")


```

## CONCLUSIONS
 * the easiest way to remove AC from the data is by aggregation.  
 * data aggregation from binary (interaction 1/0) to proportions leads to an unusual distribution of the response variable.
 * proportion data can be modelled by a glmm with beta distribution. 
 * a glmm with beta distribution fits our data better than a Gaussian model as shown by both posterior predictive checks and the fit between the predicted line and the raw data.
